---
title: "Neighborhood Poverty vs. Family Income as Predictors of Academic Outcomes"
output: html_document
date: "2026-04-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I consulted Claude.io for information about how to write this code. Lines 254-283 of this code file were copied from claude to create the interaction plots. I also used it for 191 to 211 to combine the posterior graphs of 

https://stackoverflow.com/questions/bayesian-inference-in-R was consulted for information (Be sure to include enough information that I can find the page you looked at!).

# Research Question:Neighborhood Poverty vs. Family Income as Predictors of Academic Outcomes


### Data Ingestion
```{r}
library(tidyverse)
test_scores<- read_csv("seda_geodist_pool_cs_6.0.csv") 

socioeconomics<- read_csv("seda_cov_geodist_pool_6.0.csv")
```

```{r}
#Merge using an inner join.

merged<- inner_join(test_scores, socioeconomics, by = "sedalea")


#Subseting dataset to use the variables i need
#For all students

df_all <- merged %>%
  filter(subgroup == "all") %>%
  select(sedalea, sedaleaname.x, cs_mn_avg_ol,
         lninc50avgall, povertyavgall, urbanicity) %>%
  rename(distname     = sedaleaname.x,
         test_score   = cs_mn_avg_ol,
         fam_income   = lninc50avgall,
         nbhd_poverty = povertyavgall) %>%
  drop_na(test_score, fam_income, nbhd_poverty, urbanicity) %>%
  mutate(urbanicity = factor(urbanicity,
                             levels = c("City", "Suburb", "Town", "Rural")))


table(df_all$urbanicity)
summary(df_all)
```


### Univariate Analysis for All students
```{r}
library(ggplot2)
#Test Scores
hist(df_all$test_score, main = paste("Distribution of Average Test Scores by District"), xlab = "Meant Test Score (Cohort Scale)", ylab = "Number of Districts", col = "lightyellow")

#Family Income
hist(df_all$fam_income, main = paste("Distiribution of Median Household Income by District"), xlab = "Log Median Family Income Averaged over all Students", ylabcol ="lightblue2")

#Neighborhood Poverty Rate
hist(df_all$nbhd_poverty, main = paste("Distribution of Child Poverty over Districts"), xlab = "Proportion of Children in Poverty", ylab = "Number of Districts", col="pink")

#Test scores over Neighborhood type
ggplot(df_all, aes(x = urbanicity, fill = urbanicity)) +
  geom_bar(color = "white") +
  scale_fill_manual(values = c("City"   = "steelblue",
                               "Suburb" = "darkseagreen",
                               "Town"   = "goldenrod",
                               "Rural"  = "firebrick")) +
  labs(title = "Number of Districts by Urbanicity",
       x = "District Type",
       y = "Number of Districts") +
  theme(legend.position = "none")

```


### Bivariate Analysis All Students
```{r}
library(ggplot2)

#Family Income vs Test Scores
ggplot(df_all, aes(x = fam_income, y = test_score)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", color = "orchid2", se = TRUE) +
  labs(title = "Family Income vs. Test Scores",
       subtitle = "The conventional story",
       x = "Log Median Household Income",
       y = "Mean Test Score")



# Neighborhood Poverty vs Test Scores 
ggplot(df_all, aes(x = nbhd_poverty, y = test_score)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", color = "orangered2", se = TRUE) +
  labs(title = "Neighborhood Poverty Rate vs. Test Scores",
       subtitle = "The neighborhood story",
       x = "Share of Children in Poverty",
       y = "Mean Test Score")


#checking for collinearity
#Family Income vs Neighborhood Poverty
ggplot(df_all, aes(x = fam_income, y = nbhd_poverty)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", color = "chartreuse", se = TRUE) +
  labs(title = "Family Income vs Neighborhood Poverty",
       subtitle = "How Correlated are the two predictors",
       x = "Log Median Household Income",
       y = "Proportion of Chidren in Poverty")


#Test scores by urbanicity 
ggplot(df_all, aes(x = urbanicity, y = test_score, fill = urbanicity)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("City"   = "steelblue",
                               "Suburb" = "darkseagreen",
                               "Town"   = "goldenrod",
                               "Rural"  = "firebrick")) +
  labs(title = "Test Scores by Urbanicity",
       subtitle = "How does district type relate to academic performance?",
       x = "District Type",
       y = "Mean Test Score") +
  theme(legend.position = "none")


#Neighborhood poverty by urbanicity
ggplot(df_all, aes(x = urbanicity, y = nbhd_poverty, fill = urbanicity)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("City"   = "steelblue",
                               "Suburb" = "darkseagreen",
                               "Town"   = "goldenrod",
                               "Rural"  = "firebrick")) +
  labs(title = "Neighborhood Poverty by Urbanicity",
       subtitle = "Do poverty rates differ across district types?",
       x = "District Type",
       y = "Share of Children in Poverty") +
  theme(legend.position = "none")



```

## Frequentist Analysis

```{r}
#Correlation plots - all students
cor.test(df_all$fam_income,   df_all$test_score)
cor.test(df_all$nbhd_poverty, df_all$test_score)

```

###Regression Models
```{r}
library(car)
#All students
model1<- lm(test_score ~ fam_income + nbhd_poverty + urbanicity,data = df_all)

vif(model1)
summary(model1)

```

## Bayesian Method
```{r}

#For all students
library(BayesFactor)
lmBF<- lmBF(test_score ~ fam_income + nbhd_poverty + urbanicity, data=df_all, posterior = TRUE, iterations=10000)


summary(lmBF)
par(mar = c(2, 2, 2, 2))
plot(lmBF)

```


### #Posterior Distribution Plots
```{r}

post_all    <- as.data.frame(lmBF)

colnames(post_all)
```
```{r}
# ── COMBINED POSTERIOR — family income and neighborhood poverty ────────────────
post_long <- post_all %>%
  select(`fam_income-fam_income`, `nbhd_poverty-nbhd_poverty`) %>%
  pivot_longer(cols = everything(),
               names_to  = "predictor",
               values_to = "value") %>%
  mutate(predictor = case_when(
    predictor == "fam_income-fam_income"     ~ "Family Income",
    predictor == "nbhd_poverty-nbhd_poverty" ~ "Neighborhood Poverty"
  ))

ggplot(post_long, aes(x = value, fill = predictor)) +
  geom_density(alpha = 0.6, color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_fill_manual(values = c("Family Income"        = "steelblue",
                               "Neighborhood Poverty" = "firebrick")) +
  facet_wrap(~ predictor, scales = "free_x") +
  labs(title    = "Posterior Distributions of Predictor Effects",
       subtitle = "Dashed line = zero effect",
       x = "Coefficient Value",
       y = "Density") +
  theme(legend.position = "none")

# ── FAMILY INCOME POSTERIOR ───────────────────────────────────────────────────
ggplot(post_all, aes(x = `fam_income-fam_income`)) +
  geom_density(fill = "steelblue", alpha = 0.6, color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_vline(xintercept = mean(post_all$`fam_income-fam_income`),
             color = "firebrick", linewidth = 1) +
  labs(title    = "Posterior Distribution — Family Income Effect",
       subtitle = "Red line = posterior mean, dashed = zero",
       x = "Coefficient Value",
       y = "Density")

# ── NEIGHBORHOOD POVERTY POSTERIOR ────────────────────────────────────────────
ggplot(post_all, aes(x = `nbhd_poverty-nbhd_poverty`)) +
  geom_density(fill = "firebrick", alpha = 0.6, color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_vline(xintercept = mean(post_all$`nbhd_poverty-nbhd_poverty`),
             color = "steelblue", linewidth = 1) +
  labs(title    = "Posterior Distribution — Neighborhood Poverty Effect",
       subtitle = "Red line = posterior mean, dashed = zero",
       x = "Coefficient Value",
       y = "Density")


# ── CREDIBLE INTERVALS ────────────────────────────────────────────────────────
ci_all    <- quantile(post_all$`nbhd_poverty-nbhd_poverty`,
                      probs = c(0.025, 0.975))

```

#### Supplemental Analysis about Urbanicity and 
```{r}
df_all <- df_all %>%
  mutate(income_group = case_when(
    fam_income < quantile(fam_income, 0.33) ~ "Low Income",
    fam_income < quantile(fam_income, 0.67) ~ "Middle Income",
    TRUE                                     ~ "High Income"
  ) %>% factor(levels = c("High Income", "Middle Income", "Low Income")))

```

#### Interaction Plots
```{r}
#Income group and neighborhood poverty
ggplot(df_all, aes(x = nbhd_poverty, y = test_score, color = income_group)) +
  geom_point(alpha = 0.2, size = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c("Low Income"    = "firebrick",
                                "Middle Income" = "goldenrod",
                                "High Income"   = "steelblue")) +
  labs(title    = "Does Neighborhood Poverty Hurt Low-Income Districts More?",
       subtitle = "Steeper slope for low income = interaction effect",
       x = "Share of Children in Poverty",
       y = "Mean Test Score",
       color = "Income Group")
```
```{r}
#Urbanicity and neighborhood poverty

ggplot(df_all, aes(x = nbhd_poverty, y = test_score, color = urbanicity)) +
  geom_point(alpha = 0.2, size = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c("City"   = "steelblue",
                                "Suburb" = "darkseagreen",
                                "Town"   = "goldenrod",
                                "Rural"  = "firebrick")) +
  labs(title    = "Neighborhood Poverty vs. Test Scores by District Type",
       subtitle = "Different slopes = interaction effect",
       x = "Share of Children in Poverty",
       y = "Mean Test Score",
       color = "District Type")
```

#### Frequentist Interaction Models

```{r}
# Interaction 1: poverty and income group
model2<- lm(test_score ~ fam_income + nbhd_poverty * income_group,data = df_all)
summary(model2)

# Interaction 2: poverty and urbanicity
freq_int_urban <- lm(test_score ~ fam_income + nbhd_poverty * urbanicity, data = df_all)
summary(freq_int_urban)

```

#### Bayesian Interaction Models

```{r}
# Interaction 1: poverty and income group
lmBF2 <- lmBF(
  test_score ~ fam_income + nbhd_poverty + income_group + nbhd_poverty:income_group,
  data       = as.data.frame(df_all),
  posterior  = TRUE,
  iterations = 10000
)
summary(lmBF2)
par(mar = c(2, 2, 2, 2))
plot(2)


lmBF3 <- lmBF(
  test_score ~ fam_income + nbhd_poverty + urbanicity + nbhd_poverty:urbanicity,
  data       = as.data.frame(df_all),
  posterior  = TRUE,
  iterations = 10000
)
summary(lmBF3)
par(mar = c(2, 2, 2, 2))
plot(lmBF3)
```